Data Specialization for a Staged Lambda Calculus

### 2015-02-27 paper/intro.tex
intro, work in progress

Consider a function $F$ which takes an $x$ and $y_1,\dots,y_{m-1}$, in order to
compute $f(x,y_i)$ for each $y_i$:
\begin{lstlisting}
fun F(x, @$y_0$@, @$y_1$@, @$\ldots$@ @$y_{m-1}$@) = 
  f(x, @$y_0$@);  f(x, @$y_1$@); @$\ldots$@ f(x, @$y_{m-1}$@)
\end{lstlisting}
%
Such computations, in which a multivariate function is applied many times with
one argument fixed, are common. Many algorithm design techniques revolve around
the idea of finding the computations in $f$ that depend only on $x$, and
performing them only once. Similarly, compiler optimization techniques like loop
hoisting and common subexpression elimination serve to identify the computations
common to all invocations of $f$, and perform them a single time at the
beginning of the computation.

These techniques are particularly essential in real-time computer graphics; in
fact, modern graphics architectures \emph{require} that code be separated into
distinct {\em passes} based on their required frequency of execution. For
example, graphics computations are expressed as a sequence of passes where each
pass may perform a computation at increasingly finer grain, e.g., per object,
per region of the screen, and per pixel, where later computations use the
results from the prior computations.

While performant, programming with manually separated passes results
in complex code where invariants must hold across different passes,
and local changes for one pass may require changes to other passes to
``plumb'' inputs from all other frequencies\,\cite{Foley:2011}.
%
In other words, manual pass separation breaks key programming
abstractions such as composition and modularity.
%
Graphics researchers therefore have suggested using explicitly staged
programming languages\,\cite{Proudfoot:2001,Foley:2011,He:2014},
deferring to the compiler the tedious task of separating the code into
passes. However, all prior efforts have been limited to simple
languages that do not support recursion or first-class functions.

Programming-languages researchers have studied similar problems. In
the 1980s, J{\o}rring and Scherlis identified {\em frequency
reduction} and {\em precomputation}~\cite{JS86-staging} as common
mechanisms for evaluating efficiently staged computations.  The idea
behind frequency reduction is to identify computations that are
performed multiple times and hoist them so that they can be performed
once and used later as needed.  The idea behind precomputation is to
identify computations that can be performed earlier---for example, at
compile time if their inputs are known statically---and perform them
at that earlier time.

There has been much work on language-based techniques for performing
precomputation. Partial evaluation~\cite{futamura71,jones96} achieves
precomputation by specializing functions for specific values. Going back to our
example, we can specialize $f$ at $[1,2, \ldots, n]$, written $f_{[1,2, \ldots,
n]}$, such that $f(x,y) = f_{[1,2, \ldots, n]}(y)$ and specialize $F$ for the
same input as
\begin{lstlisting}
fun @$F_{[1,2, \ldots, n]}(y_0, y_1 \ldots y_{m-1})$@ = 
  @$(f_{[1,2, \ldots, n]}(y_0);  f_{0}(y_1)  \ldots f_{0}(y_{m-1}))$@.
\end{lstlisting}
%
Partial evaluation typically proceeds by performing a
binding-time analysis to determine the static and dynamic parts of $f$
and $F$ and specializes them based on the results of the binding-time
analysis. A closely related technique, {\em metaprogramming}, enables
the programmer to write programs that generate specialized code.
Metaprogramming enables fine-grained control over specialization by
requiring staging annotations that make explicit the stage of each
computation.

>>>>

Consider a function $F$ which takes an $x$ and $y_1,\dots,y_{m-1}$, and computes
$f(x,y_i)$ for each $y_i$:
\begin{lstlisting}
fun @$F(x, y_0, y_1, \ldots, y_{m-1})$@ = 
  @$f(x, y_0);  f(x, y_1); \ldots; f(x, y_{m-1})$@
\end{lstlisting}
%
This scenario, in which a multivariate function is applied many times with one
argument fixed, arises frequently. 
\crem{?? Many algorithm design techniques revolve around the idea of finding the
computations in $f$ that depend only on $x$, and
performing them only once.
\\
?? It's advantageous to find the computations in f that depend on x and perform
them only once.}
This is the intent of compiler optimization techniques like loop hoisting and
common subexpression elimination, which identify the computations common to all
invocations of $f$, and perform them once at the beginning of the computation.

Programs structured in this way are particularly important in real-time computer
graphics; in fact, modern graphics architectures \emph{require} that graphics
computations be organized into distinct {\em passes} which perform increasingly
fine-grained computations (for example, per object, per region of the screen,
and per pixel, where later computations use the results from the prior
computations).

While performant, this paradigm of programming with explicit collections of
passes results in complex code where invariants must hold across different
passes, and local changes for one pass may require changes to other passes to
``plumb'' inputs from all other frequencies\,\cite{Foley:2011}.
%
In other words, requiring manual separation of passes breaks key programming
abstractions such as composition and modularity.
%
Graphics researchers therefore have suggested using explicitly staged programs
\,\cite{Proudfoot:2001,Foley:2011,He:2014}, deferring to the compiler the
mechanical task of separating such programs into passes. However, prior work in
automatic pass separation has been limited to simple languages that do not
support recursion or first-class functions.

Programming languages researchers have studied similar problems. In the 1980s,
J{\o}rring and Scherlis identified {\em frequency reduction} and {\em
precomputation}~\cite{JS86-staging} as mechanisms for efficiently evaluating
staged computations. To perform frequency reduction, one identifies and hoists
computations that are performed multiple times, in order to compute them only
once. To perform precomputation, one identifies computations that can be
performed in advance, and does so---for example, at compile time if the relevant
inputs are statically known.

There has been much work on language-based techniques for performing
precomputation. Partial evaluation~\cite{futamura71,jones96} achieves
precomputation by specializing functions to specific arguments. Going back to
our example, if we specialize $f$ to a particular $v$, written $f_v$, such that
$f(v,y) = f_v(y)$, then we can specialize $F$ to $v$ as
\begin{lstlisting}
fun @$F_v(y_0, y_1, \ldots, y_{m-1})$@ = 
  @$f_v(y_0);  f_v(y_1); \ldots; f_v(y_{m-1})$@.
\end{lstlisting}
eliminating the need to recompute
\crem{WIP especially below here}
TODO.
%
A closely related technique, {\em metaprogramming}, allows writing programs that
generate specialized code. Metaprogramming enables fine-grained control over
specialization by requiring staging annotations that make explicit the stage of
each computation. (In partial evaluation, we typically assume that a
\emph{binding-time analysis} has automatically determined which parts of $f$ can
be specialized.)
